---
title: "Chapter 6"
date: "`r Sys.Date()`"
output: html_document
---

# Assignment 6 - Data analysis exercise

```{r}
#Load the required packages
library(tidyverse)
library(lme4)

#Load the data
BPRS <- read_csv("data/BPRS.csv")
RATS <- read_csv("data/RATS.csv")
RATS_original <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')


#Convert the categorical variables of both data sets to factors
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)


```

## Part I

We are going to analyse the RATS dataset, which contains data about the weight change of three groups of rats on different diets. Weight is measured on different days, each of which is its own variable from WD1 to WD64.

```{r}
str(RATS)
summary(RATS)
```

First we well visualize the data.

```{r}
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight)))
```

We can see that the weight of the rats seem to increase in all groups, maybe most in group 2. Overall weight is lowest in group 1.

Next we will standardize the values and redraw the plot.

```{r}
# Standardise the variable Weight
RATS <- RATS %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight-mean(Weight))/sd(Weight)) %>%
  ungroup()

# Plot again with the standardised Weight
ggplot(RATS, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")+
  theme(legend.position = "none")
```

After standardizing the differences in time look much smaller.

Next we will create some summary graphs.

```{r}
n <- 11

# Summary data with mean and standard error of weight by treatment and week 
RATSS <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), n=n(), se = sd(Weight)/sqrt(n)) %>%
  ungroup()


# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

We can see that the weight seems to increase over time in all groups. Weight is largest in group 3, then 2 and finally 1.

We will use the mean of days 8-64 as the summary measure to look how the different treatments affect the weight. We will also remove some outliers.

```{r}
# Create a summary data by group and ID with mean as the summary variable (ignoring baseline day 1)
RATS_summary <- RATS %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Draw a boxplot of the mean versus group
ggplot(RATS_summary, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")

# Create a new data by filtering the outliers and adjust the ggplot code the draw the plot again with the new data
RATS_summary_no_outliers <- RATS_summary %>% filter(mean >250 & mean <550)
ggplot(RATS_summary_no_outliers, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), days 8-64")
```

After removing the outliers the variation in the groups is significantly reduced. There is still the above mentioned differences in the weight. However, we cannot really make any conclusions with these results, as they are mainly affected by the baseline weight. We can use ANOVA to account for this.

```{r}
# Add the baseline from the original data as a new variable to the summary data
RATS_summary <- RATS_summary %>%
  mutate(baseline = RATS_original$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline+Group, data = RATS_summary)

# Compute the analysis of variance table for the fitted model
anova(fit)
```

From the ANOVA table we can see that when adjusting for the baseline value, the group is not anymore significant (p\>0.05). This means that the treatment has no effect on the weight. The baseline value is naturally highly significant.

## Part II

We are going to analyse the BPRS dataset, which contains data about psychiatric patients assigned to two different treatments. BPRS scores are measured at time 0 and 1-8 weeks.

```{r}
str(BPRS)
summary(BPRS)
```

We are going to analyse the effect of the chosen treatment on the bprs score. First we will plot the data.

```{r}
ggplot(BPRS, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)))
```

We can see that the scores seem to be decreasing over time in both groups.

We are first going to ignore the repeated-measures structure of the data, and we will fit a multiple linear regression model with bprs as response and week and treatment as explanatory variables.

```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs  ~ week +treatment , data=BPRS)

# print out a summary of the model
summary(BPRS_reg)
```

In this model week is a significant predictor but treatment is not. This means that there are significant differences in bprs scores over time, but the treatment does not play a role.

The previous model assumes independence of the repeated measures of bprs, and this assumption is highly unlikely. So, now we will move on to consider both some more appropriate models. To begin the more formal analysis of the data, we will first fit the *random intercept model* for the same two explanatory variables. Fitting a random intercept model allows the linear regression fit for each rat to differ in *intercept* from other rats.

```{r}
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

The parameters are very close to the ones from the linear model, and again only week seems to be significant.

Next we fit the *random intercept and random slope model* to the data. Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the bprs score profiles, but also the effect of time.

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

From this model we get very similar estimates on the parameters. However, the likelihood ratio test between the models is significant, meaning that the new model is a better fit.

Finally, we can fit a random intercept and slope model that allows for a treatment × week interaction.

```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week * treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

In this model the treatment is still not significant, as is not the interaction. The likelihood ratio test is not significant, meaning that this model does not improve on the previous one.

Finally we are going to plot the observed values and fitted values of the BPRS_ref1 model.

```{r}
# draw the plot of BPRS with the observed Weight values
ggplot(BPRS, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)), name = "Observed BPRS")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted to BPRS
BPRS$Fitted <- Fitted

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRS, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)), name = "Fitted BPRS")
```
