---
---
---

# Assignment 2 - Data analysis exercise

*Describe the work you have done this week and summarize your learning.*

-   Describe your work and results clearly.
-   Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
-   Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

### Data analysis

First we load the required packages and read the data.

We use the dataset learning2014 (<http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt>), which contains data about a survey of university students and their attitudes towards learning.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(GGally)
lrn14 <- read_csv("data/learning2014.csv")
```

```{r}
#Dimensions of the data
dim(lrn14)
#Structure of the data
str(lrn14)
```

The dataset contains 166 observations (students) and 7 variables: age, gender, exam points (points) and summary variables of questions related to deep learning (deep), surface learning (surf), strategic learning (stra) and students' attitude towards statistics (attitude).

```{r}
ggpairs(lrn14, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20)))
```

Next we explore the data using a scatter plot matrix. From the figure we can see the following:

-   The variables seem to be mostly normally distributed. Age is skewed right and points are skewed left
-   There are more women than men in the dataset
-   Males seem to have on average better attitude
-   There is a significant positive correlation between attitude and points
-   There is a milder negative correlation between surface and deep learning (which is to be expected)

We want to make a model from this data where we want to explain the amount of points. We are going to use a linear regression model and choose three explanatory variables. For the first model we are going to choose the demographic variables age and gender and also attitude, which had the highest correlation with points.

```{r}
model1 <- lm(points ~ age+gender+attitude, data=lrn14)
summary(model1)
```

We can see that from these variables only attitude was a significant predictor, because it had a p-value \<0.05. Age and gender were not significantly associated with points in this model. In order to try to improve the model we replace age and gender with the two variables that had the highest correlation with points besides attitude: stra and surf.

```{r}
model2 <- lm(points ~ attitude+stra+surf, data=lrn14)
summary(model2)
```

Stra and surf were neither associated with points, thus we are left with a model with only attitude as the explaining variable:

```{r}
model3 <- lm(points ~ attitude, data=lrn14)
summary(model3)
```

In this model attitude is highly significantly associated with points (p=4.12e-09). The estimate for attitude is 3.5 meaning that for every one point increase in attitude the points will increase by 3.5. The intercept is 11.6, meaning that if the attitude is 0 the points will be 11.6. For this model the R-squared is 0.19 meaning that this model explains about 19% of the variance in points.

To analyse whether the assumptions of linear regression are met, we are going to draw the diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage.

```{r}
par(mfrow = c(2,2))
plot(model3, which=c(1,2,5))
```

The main assumptions of linear regression are

-   Linear relationship between predictors and outcome
-   The residuals are independent of each other
-   The variance of residuals is equal and does not depend on the value of the response variable
-   The residuals are normally distributed

The linear relationship between attitude and points can be seen from the previous scatter plot. From the residuals vs. fitted plot we can see that residuals are evenly spaced throughout the fitted values, meaning that the assumption of equal variance of residuals is met and they seem to be independent of each other. From the Q-Q plot we can see that the residuals follow quite well a straight line (except for some differences in the extremes), meaning that the assumption of normality of the residuals is met. Finally, from the residuals vs. leverage plot we can see that none of the points is close to the Cook's distance, meaning that there are no outliers which would influence the model too strongly.
